|

### 2023-02-05 manuscript/manuscript.Rmd
more updates

\begin{abstract}
Patient reported outcomes including quality of life (QoL) assessments are increasingly being included as either primary or secondary outcomes in randomized controlled trials. While making the outcomes more relevant for patients it entails a challenge in cases where death or a similar event makes the outcome of interest undefined. A pragmatic - and much used - solution is to assign diseased patient with the lowest possible QoL score. This makes medical sense, but creates a statistical problem since traditional tests such as t-tests or Wilcoxon tests potentially loses high amounts of statistical power. In this paper we propose a novel test that can keep the medically relevant composite outcome but preserve full statistical power. The test is also applicable in other situations where a specific value (say zero days alive outside hospitals) encodes a special meaning. The test is implemented in an R package which is available for download. The paper includes a simulation study and an application to a COVID-19 trial.
\end{abstract}

@@

In medical intervention research including in particular randomized controlled trials (RCTs) there is a trend towards increased use of patient reported outcomes; Quality of Life (QoL) scores being one of the most prominent of these. Established statistical practice is to compare these between treatment groups using non-parametric tests such as Mann–Whitney–Wilcoxon rank-sum test (@wilcoxon3001968individual, @mann1947test), hence henceforth Wilcoxon, since distributions are rarely normal or symmetric. When all patients are alive and able and willing to provide QoL scores at the scheduled measurement time this procedure works very well. Not all clinical settings, however, will have all participants alive at time of scheduled assessment of QoL. This is particularly true in trials within intensive care where mortality of approximately 30% is not uncommon and which is the setting that motivated this work. A pragmatic solution which is gaining popularity is to simply give the deceased patients the lowest possible score and then use a Wilcoxon test or similar to compare groups. This paper will demonstrate that this approach can lead to dramatic loss of statistical power. The paper will also present an alternative and novel statistical test that avoids the power loss. The novel test procedure is implemented in R and made publicly available.

The key-insight in the proposed test is to incorporate that the outcome is actually a two-dimensional outcome of a very special type, and that the constructed combined outcome follows a continuous-singular mixture distribution. This unusual distribution is why one cannot resort to non-parametric Wilcoxon tests since the singular component in the distribution of the combined outcome will get reduced to simple ties. It is noted that the handling of ties in standard statistical software varies and is opaque. However, the handling of ties is not the main reason why the Wilcoxon test suffers power loss. The main reason is that the null-hypothesis in these Wilcoxon type tests (stochastic domination) does not handle the empirical fact that treatments might influence mortality and QoL differently.

As an alternative we propose to model the binary component (i.e., survival) and the continuous part (i.e., actual QoL) separately but to construct a single test for no treatment effect on either based on a likelihood ratio. We can thus provide a single p-value for the hypothesis of no treatment effect on the extended QoL where death is given the lowest possible score. To accommodate potential non-normality of the recorded QoL scores we consider both a parametric and a semi-parametric version of the test where the semi-parametric version is as widely applicable as the Wilcoxon test. Both test procedures will provide effect estimates of mean differences between treatment groups based on the combined outcome along with confidence intervals. This is also an added benefit compared to the Wilcoxon-type testing approach.

It should be noted that while we in this paper exemplify the procedure using QoL and mortality the method is applicable in any setting where a single value of a combined outcome has a special interpretation compared to an otherwise continuous scale. Another example from intensive care research consider the outcome "days alive and out of hospital within 90 days from randomization". Here in-hospital fatalities will all have the value 0, while everybody else will have outcomes ranging from 0 to 90. Such outcomes are also routinely analyzed using Wilcoxon-type tests. Our proposed test would increase power while also providing a mean effect estimate along with confidence interval. Note also that the developed mathematical method allows straightforwardly for the inclusion of confounding variables or other adjustment factors. The method itself is therefore just as applicable in non-randomized studies or epidemiological studies in general.

@@

We propose to test the null-hypothesis of no treatment effect on the combined outcome by a likelihood ratio test of the joint distribution of $Y_i \mid A_i$ and $A_i$. This has the advantage that it increases the efficiency compared the Wilcoxon test and it yields a single p-value appropriate for testing a primary outcome in a clinical trial. Let $(Y_i, A_i, R_i, X_i)$ for $i = 1, \ldots, n$ be independent and identically distributed random variables. We can write our model in a general form as 
\begin{align*}
  Y_i \mid A_i = 1, R_i, X_i &\sim F(Y_i; \mu_i(R_i, X_i), \Psi)\\
  A_i \mid R_i, X_i &\sim \text{Bernoulli}(A_i; \pi_i(R_i, X_i))  
\end{align*}
for some mean functions $\pi_i$ and $\mu_i$ and a distribution function $F$ characterizing the distribution of the observed continuous outcomes with possible nuisance parameter $\Psi$. The joint likelihood function for the combined outcome conditional on treatment and baseline covariates is
\begin{align*}
 L_n(\mu_i(R_i, X_i), \pi_i(R_i, X_i), \Psi) &= \prod_{i=1}^n f(Y_i; \mu_i(R_i, X_i), \Psi)^{A_i}P(A_i = a_i; \pi_i(R_i, X_i))\\
     &= \prod_{i=1}^n f(Y_i; \mu_i(R_i, X_i), \Psi)^{A_i} \pi_i(R_i, X_i)^{A_i} (1 - \pi_i(R_i, X_i))^{1-A_i}\nonumber
\end{align*}
We note that when $F$ admits an absolutely continuous density function the likelihood function can equivalently be written solely in terms of the combined outcome $\widetilde{Y}$ in Equation (\ref{eq:combinedOutcome}) since $f(Y_i; \cdot)^{A_i} = f(\widetilde{Y}_i; \cdot)^{1_{\widetilde{Y}_i \ne \mathcal{E}}}$ and $\pi_i(\cdot)^{A_i} = \pi_i(\cdot)^{1_{\widetilde{Y}_i \ne \mathcal{E}}}$ almost surely. An important property of this model is that the likelihood function factorizes into two components as
\begin{align}
 L_n(\mu_i(R_i, X_i), \pi_i(R_i, X_i)) &= L_{1,n}(\mu_i(R_i, X_i), \Psi) L_{2,n}(\pi_i(R_i, X_i))\label{eq:likFactorization}
\end{align}
where $L_{1,n}(\mu_i(R_i, X_i), \Psi) = \prod_{i=1}^n f(Y_i; \mu_i(R_i, X_i), \Psi)^{A_i}$ and $L_{2,n}(\pi_i(R_i, X_i)) = \prod_{i=1}^n \pi_i(R_i, X_i)^{A_i} (1 - \pi_i(R_i, X_i))^{1-A_i}$. This implies that the parameters for the observed outcomes, $\mu_i(R_i, X_i)$, can be estimated independently of the parameters governing the probability of observing the outcome, $\pi_i(R_i, X_i)$. This factorization is a consequence of the likelihood construction and it does neither assume nor require independence between the value of the outcome and the probability of observing it.

@@

 use the profile likelihood ratio test (LRT) statistic which can be written as a function of the treatment effects as
and the value of the LRT statistic under the null-hypothesis of no treatment effect is therefore $W_n(0, 0)$. Similar to the factorization of the likelihood function in Equation (\ref{eq:likFactorization}), the LRT statistic in Equation (\ref{eq:lrtgeneral})  also decomposes into the sum
\begin{align}
  W_n(\mu_\delta, \beta_\delta) &= W_{1,n}(\mu_\delta) + W_{2,n}(\beta_\delta)\label{eq:LRTsum} 
\end{align}
of two LRT statistics -- one for the continuous part and one for the discrete part of the combined outcome. Under very general conditions it follows that $W_n(\mu_\delta, \beta_\delta)$ is approximately $\chi^2$ distributed with two degrees of freedom [@wilks1938large; @owen1988empirical].

@@

With the generalized linear models in Equations (\ref{eq:glm1}) and (\ref{eq:glm2}) the first term is the LRT statistic with a single degree of freedom in a linear regression model, and the second term is the LRT statistic with a single degree of freedom in a logistic regression model. This makes this test very easy to perform in standard statistical software that outputs the likelihood value for a model fit without the need of special methods. The calculation of the LRT simply amounts to estimating two linear regression models and two logistic regression models -- for each type a model with and a model without the binary treatment indicator as a covariate. Combining the two likelihood ratios according to Equation (\ref{eq:LRTsum}) calculates our test statistic, and the p-value for no treatment effect can be determined through the $\chi^2$-distribution with two degrees of freedom. The critical value for rejecting the null-hypothesis of no treatment effect at the 5% level is equal to $5.99$ and equal to $9.21$ at the 1% level.

@@

A drawback of the parametric LRT introduced in the previous section is that it requires deciding on a parametric distribution for the observed outcomes. In this section we introduce a more flexible approach based on an empirical LRT. This approach also utilizes the additive decomposition of the LRT statistic in Equation (\ref{eq:LRTsum}) but substitutes the term $W_{1,n}$ with a term that is free of any distributional assumptions. Combining this with the binomial model for $A$ leads to a semi-empirical LRT for the treatment effect.

Let $\mathcal{I}_0 = \left\{i = 1,\ldots, n : R_i = 0, A_i = 1\right\}$ and $\mathcal{I}_1 = \left\{i = 1,\ldots, n : R_i = 1, A_i = 1\right\}$ be the sets of indices of the observed outcomes for the two treatments. The empirical LRT statistic as a function of the difference in expected value is given by
\begin{align*}
  W_{1,n}^E(\mu_\delta^\ast) &= 2\sup_{\mu}\left(\sum_{i \in \mathcal{I}_0}\log \left(1 + \lambda_1 (Y_i - \mu)\right) + \sum_{j \in \mathcal{I}_1}\log\left(1 + \lambda_2 (Y_j - \mu - \mu_\delta^\ast)\right)\right)
\end{align*}
where $\lambda_1$ and $\lambda_2$ are the solutions to the following equations
\begin{align*}
  |\mathcal{I}_0|^{-1}\sum_{i \in \mathcal{I}_0} \frac{Y_i - \mu}{1 + \lambda_1 (Y_i - \mu)} = 0, \quad |\mathcal{I}_1|^{-1}\sum_{j \in \mathcal{I}_1}\frac{Y_j - \mu - \mu_\delta^\ast}{1 + \lambda_2 (Y_j - \mu - \beta_\delta^\ast)} = 0
\end{align*}
We refer to the appendix for a derivation of the test statistics.

The semi-parametric LRT statistic for testing the null-hypothesis of no treatment effect is equal to $W_n^{SP}(0, 0)$ where 
\begin{align}
  W_n^{SP}(\mu_\delta^\ast, \alpha_\delta^\ast) = W_{1,n}^E(\mu_\delta^\ast) + W_{2,n}(\alpha_\delta^\ast) \label{eq:LRT_SP_total}
\end{align}
By the non-parametric Wilk's theorem [@owen1988empirical] it follows that $W_n^{SP}(0, 0) \sim \chi^2_2$ asymptotically.

@@

To assess the power we vary the sample size between 50 to 350 observations in each group with increments of 25, and for each scenario we perform 25,000 simulations and estimate the power by the proportion of rejected null hypotheses at the 5\% level. We compare the power of both our parametric and semi-parametric likelihood ratio tests to the two-sample t-test and the Wilcoxon test. Figure \ref{fig:powerCurves} shows the estimated power functions. The results are also presented in table form in the supplementary material.

@@

In addition to the power simulations we also performed a simulation study for the type I error. We considered two scenarios similar to scenarios 1 and 4 in Table \ref{tab:simulation} under a null-hypothesis of no treatment effect. Specifically, we considered $A_i \sim \Bernoulli(\pi)$, $\widetilde{Y}_i \mid A_i = 1 \sim N(3,1)$ and $A_i \sim \Bernoulli(\pi)$, $\widetilde{Y}_i \mid A_i = 1 \sim \Beta(1, 1)$. We estimated the type I error through simulation for different sample sizes and $\pi \in \left\{0.2, 0.4, 0.6, 0.8\right\}$. The results are shown in the supplementary material.

@@

The COVID-STEROID2 trial was a multi-center study comparing daily dose 6 mg (low) of dexamethasone for up to 10 days vs. 12 mg (high) in patients with severe and critical COVID-19. The full description of the study including all results can be found in @10.1001/jama.2021.18295. The primary outcome of the trial was days alive and without out use of life-support in the period from randomization to day 28. In other words, that outcome is an integer between 0 and 28 with a substantial peak at day 0 coming from the patients who are never taken off life-support systems. The analyses in the main publications (@10.1001/jama.2021.18295) were conducted using the test proposed in this paper. In the JAMA publication the test is being referred to as the Kryger Jensen and Lange test. In this section we re-analyze the sub-group consisting of patients in need of invasive mechanical ventilation at baseline (n = 206). These can generally be taken to be the most severely ill patients.

@@

```{r, echo = FALSE, fig.width = 8, fig.height = 3, fig.cap="\\label{fig:application}Histograms of days alive without life support from the The COVID-STEROID2 trial stratified by treatment dosis (left, middle) and simultaneous confidence regions based on the semi-parametric likelihood ratio test (right).", fig.pos="htbp"}
par(mfrow=c(1,3), mgp=c(2,1,0), mar=c(3, 3, 2, 0))
hist(subset(d, !d$allocation)$dawols28, 10, xlim=c(0,30),
     main="High dose", ylim = c(0,55), col="gray70", border="gray90",
     cex.axis=1.2, cex.main=1.5, cex.lab=1.3, ylab="",
     xlab="Days alive without life support")
hist(subset(d, d$allocation)$dawols28, xlim=c(0,30),
     main="Low dose", ylim=c(0,55), 10, col="gray70", border="gray90",
     cex.axis=1.2, cex.main=1.5, cex.lab=1.3, ylab="",
     xlab="Days alive without life support")
image(sCI$muDelta, sCI$logORdelta, sCI$surface,
     col = rev(fields::tim.colors(256)), useRaster = TRUE,
     xlab="Difference in means",
     ylab="log OR",
     cex.axis=1.2, cex.main=1.5, cex.lab=1.3)
points(m_SPLRT$muDelta, log(m_SPLRT$alphaDelta), pch = 19, cex = 1.5)
contour(sCI$muDelta, sCI$logORdelta, sCI$surface, add = TRUE, 
        levels = stats::qchisq(0.99, 2), lwd = 1, labels = 0.99, lty=1)
contour(sCI$muDelta, sCI$logORdelta, sCI$surface, add = TRUE, 
        levels = stats::qchisq(0.95, 2), lwd = 1, labels = 0.95, lty=1)
contour(sCI$muDelta, sCI$logORdelta, sCI$surface, add = TRUE, 
        levels = stats::qchisq(0.8, 2), lwd = 1, labels = 0.8, lty=1)
contour(sCI$muDelta, sCI$logORdelta, sCI$surface, add = TRUE, 
        levels = stats::qchisq(0.5, 2), lwd = 1, labels = 0.5, lty=1)
title("Confidence regions", cex.main=1.5)
abline(v = 0, lty=2)
abline(h = 0, lty=2)
```

Comparing the groups by a Wilcoxon test yields a p-value of $0.242$ hence non-significant at the pre-specified 5\% level. Using our proposed parametric likelihood ratio test yields a p-value of $0.019$ and thus leading to a rejection of the null-hypothesis. From the histograms it could be questioned whether the continuous part of the distributions are normal distributed. Accordingly, we also apply our proposed semi-parametric version of the likelihood ratio test. This yields a p-value of $0.021$. In conclusion, our proposed method is able to detect the difference between the groups. Further and in contrast to the Wilcoxon test, our method also provides a way to interpret the difference between the groups. This is illustrated in the right panel of Figure \ref{fig:application} showing simultaneous confidence regions for the two effect parameters. In this case it is clear that the primary effect of the intervention is not on the discrete component (i.e., the proportion of patients who never get off life support). Instead, the effect appears to be on the continuous part and is between half a day and seven days.

@@

>>>>

\begin{abstract}
Patient reported outcomes including quality of life assessments are increasingly being included as either primary or secondary outcomes in randomized controlled trials. While making the outcomes more relevant for patients it entails a challenge in cases where death or a similar event makes the outcome of interest undefined. A pragmatic - and much used - solution is to assign diseased patient with the lowest possible quality of life score. This makes medical sense, but creates a statistical problem since traditional tests such as the t-test or the Wilcoxon test potentially loses high amounts of statistical power. In this paper we propose a novel test that can keep the medically relevant composite outcome but preserve full statistical power. The test is also applicable in other situations where a specific value (say zero days alive outside hospitals) encodes a special meaning. The test is implemented in an R package. The paper includes a simulation study and an application to a COVID-19 randomized controlled trial.
\end{abstract}

@@

In medical intervention research including in particular randomized controlled trials (RCTs) there is a trend towards increased use of patient reported outcomes; Quality of Life scores being one of the most prominent of these. Established statistical practice is to compare these between treatment groups using non-parametric tests such as Mann–Whitney–Wilcoxon rank-sum test (@wilcoxon3001968individual, @mann1947test), hence henceforth Wilcoxon, since distributions are rarely normal or symmetric. When all patients are alive and able and willing to provide quality of life scores at the scheduled measurement time this procedure works very well. Not all clinical settings, however, will have all participants alive at time of scheduled assessment of quality of life This is particularly true in trials within intensive care where mortality of approximately 30% is not uncommon and which is the setting that motivated this work. A pragmatic solution which is gaining popularity is to simply give the deceased patients the lowest possible score and then use a Wilcoxon test or similar to compare groups. This paper will demonstrate that this approach can lead to dramatic loss of statistical power. The paper will also present an alternative and novel statistical test that avoids the power loss. The novel test procedure is implemented in R and made publicly available.

The key-insight in the proposed test is to incorporate that the outcome is actually a two-dimensional outcome of a very special type, and that the constructed combined outcome follows a continuous-singular mixture distribution. This unusual distribution is why one cannot resort to non-parametric Wilcoxon tests since the singular component in the distribution of the combined outcome will get reduced to simple ties. It is noted that the handling of ties in standard statistical software varies and is opaque. However, the handling of ties is not the main reason why the Wilcoxon test suffers power loss. The main reason is that the null-hypothesis in these Wilcoxon type tests (stochastic domination) does not handle the empirical fact that treatments might influence mortality and quality of life differently.

As an alternative we propose to model the binary component (i.e., survival) and the continuous part (i.e., actual quality of life) separately but to construct a single test for no treatment effect on either based on a likelihood ratio. We can thus provide a single p-value for the hypothesis of no treatment effect on the extended quality of life where death is given the lowest possible score. To accommodate potential non-normality of the recorded quality of life scores we consider both a parametric and a semi-parametric version of the test where the semi-parametric version is as widely applicable as the Wilcoxon test. Both test procedures will provide effect estimates of mean differences between treatment groups based on the combined outcome along with confidence intervals. This is also an added benefit compared to the Wilcoxon-type testing approach.

It should be noted that while we in this paper exemplify the procedure using quality of life and mortality the method is applicable in any setting where a single value of a combined outcome has a special interpretation compared to an otherwise continuous scale. Another example from intensive care research consider the outcome "days alive and out of hospital within 90 days from randomization". Here in-hospital fatalities will all have the value 0, while everybody else will have outcomes ranging from 0 to 90. Such outcomes are also routinely analyzed using Wilcoxon-type tests. Our proposed test would increase power while also providing a mean effect estimate along with confidence interval. Note also that the developed mathematical method allows straightforwardly for the inclusion of confounding variables or other adjustment factors. The method itself is therefore just as applicable in non-randomized studies or epidemiological studies in general.

@@

We propose to test the null-hypothesis of no treatment effect on the combined outcome by a likelihood ratio test of the joint distribution of $Y_i \mid A_i$ and $A_i$. This has the advantage that it increases the efficiency compared the Wilcoxon test and it yields a single p-value appropriate for testing a primary outcome in a clinical trial. Let $(Y_i, A_i, R_i, X_i)$ for $i = 1, \ldots, n$ be independent and identically distributed random variables. We can write our model in a general form as 
\begin{align*}
  Y_i \mid A_i = 1, R_i, X_i &\sim F(Y_i; \mu_i(R_i, X_i), \Psi)\\
  A_i \mid R_i, X_i &\sim \text{Bernoulli}(A_i; \pi_i(R_i, X_i))  
\end{align*}
for some mean functions $\pi_i$ and $\mu_i$ and a distribution function $F$ characterizing the distribution of the observed continuous outcomes with possible nuisance parameter $\Psi$. The joint likelihood function for the combined outcome conditional on treatment and baseline covariates is
\begin{align*}
 L_n(\mu_i(R_i, X_i), \pi_i(R_i, X_i), \Psi) &= \prod_{i=1}^n f(Y_i; \mu_i(R_i, X_i), \Psi)^{A_i}P(A_i = a_i; \pi_i(R_i, X_i))\\
     &= \prod_{i=1}^n f(Y_i; \mu_i(R_i, X_i), \Psi)^{A_i} \pi_i(R_i, X_i)^{A_i} (1 - \pi_i(R_i, X_i))^{1-A_i}\nonumber
\end{align*}
We note that when $F$ admits an absolutely continuous density function the likelihood function can equivalently be written solely in terms of the combined outcome $\widetilde{Y}$ in Equation (\ref{eq:combinedOutcome}) since $f(Y_i; \cdot)^{A_i} = f(\widetilde{Y}_i; \cdot)^{\mathbbm{1}(\widetilde{Y}_i \ne \mathcal{E})}$ and $\pi_i(\cdot)^{A_i} = \pi_i(\cdot)^{\mathbbm{1}(\widetilde{Y}_i \ne \mathcal{E})}$ almost surely. An important property of this model is that the likelihood function then factorizes into two components as
\begin{align}
 L_n(\mu_i(R_i, X_i), \pi_i(R_i, X_i)) &= L_{1,n}(\mu_i(R_i, X_i), \Psi) L_{2,n}(\pi_i(R_i, X_i))\label{eq:likFactorization}
\end{align}
where $L_{1,n}(\mu_i(R_i, X_i), \Psi) = \prod_{i=1}^n f(Y_i; \mu_i(R_i, X_i), \Psi)^{A_i}$ and $L_{2,n}(\pi_i(R_i, X_i)) = \prod_{i=1}^n \pi_i(R_i, X_i)^{A_i} (1 - \pi_i(R_i, X_i))^{1-A_i}$. This implies that the parameters for the observed outcomes, $\mu_i(R_i, X_i)$, can be estimated independently of the parameters governing the probability of observing the outcome, $\pi_i(R_i, X_i)$. This factorization is a consequence of the likelihood construction and it does neither assume nor require independence between the value of the outcome and the probability of observing it.

@@

 use the profile likelihood ratio test statistic which can be written as a function of the treatment effects as
and the value of the likelihood ratio test statistic under the null-hypothesis of no treatment effect is therefore $W_n(0, 0)$. Similar to the factorization of the likelihood function in Equation (\ref{eq:likFactorization}), the likelihood ratio test statistic in Equation (\ref{eq:lrtgeneral})  also decomposes into the sum
\begin{align}
  W_n(\mu_\delta, \beta_\delta) &= W_{1,n}(\mu_\delta) + W_{2,n}(\beta_\delta)\label{eq:LRTsum} 
\end{align}
of two likelihood ratio test statistics -- one for the continuous part and one for the discrete part of the combined outcome. Under very general conditions it follows that $W_n(\mu_\delta, \beta_\delta)$ is approximately $\chi^2$ distributed with two degrees of freedom [@wilks1938large; @owen1988empirical].

@@

With the generalized linear models in Equations (\ref{eq:glm1}) and (\ref{eq:glm2}) the first term is the LRT statistic with a single degree of freedom in a linear regression model, and the second term is the LRT statistic with a single degree of freedom in a logistic regression model. This makes this test very easy to perform in standard statistical software that outputs the likelihood value for a model fit without the need of special methods. The calculation of the likelihood ratio test simply amounts to estimating two linear regression models and two logistic regression models -- for each type a model with and a model without the binary treatment indicator as a covariate. Combining the two likelihood ratios according to Equation (\ref{eq:LRTsum}) calculates our test statistic, and the p-value for no treatment effect can be determined through the $\chi^2$-distribution with two degrees of freedom. The critical value for rejecting the null-hypothesis of no treatment effect at the 5% level is equal to $5.99$ and equal to $9.21$ at the 1% level.

@@

A potential drawback of the parametric likelihood ratio test introduced in the previous section is that it requires deciding on a parametric distribution for the observed outcomes. In this section we introduce a more flexible approach based on an empirical likelihood ratio test. This approach also utilizes the additive decomposition of the likelihood ratio test statistic in Equation (\ref{eq:LRTsum}) but substitutes the term $W_{1,n}$ with a term that is free of any distributional assumptions. Combining this with the binomial model for $A$ leads to a semi-parametric likelihood ratio test for the overall treatment effect.

Let $\mathcal{I}_0 = \left\{i = 1,\ldots, n : R_i = 0, A_i = 1\right\}$ and $\mathcal{I}_1 = \left\{i = 1,\ldots, n : R_i = 1, A_i = 1\right\}$ be the sets of indices of the observed outcomes for the two treatments. The empirical likelihood ratio test statistic as a function of the difference in expected value is given by
\begin{align*}
  W_{1,n}^E(\mu_\delta^\ast) &= 2\sup_{\mu}\left(\sum_{i \in \mathcal{I}_0}\log \left(1 + \lambda_1 (Y_i - \mu)\right) + \sum_{j \in \mathcal{I}_1}\log\left(1 + \lambda_2 (Y_j - \mu - \mu_\delta^\ast)\right)\right)
\end{align*}
where $\lambda_1$ and $\lambda_2$ are the solutions to the following equations
\begin{align*}
  |\mathcal{I}_0|^{-1}\sum_{i \in \mathcal{I}_0} \frac{Y_i - \mu}{1 + \lambda_1 (Y_i - \mu)} = 0, \quad |\mathcal{I}_1|^{-1}\sum_{j \in \mathcal{I}_1}\frac{Y_j - \mu - \mu_\delta^\ast}{1 + \lambda_2 (Y_j - \mu - \beta_\delta^\ast)} = 0
\end{align*}
We refer to the appendix for a derivation of the test statistics.

The semi-parametric likelihood ratio test statistic for testing the null-hypothesis of no treatment effect is equal to $W_n^{SP}(0, 0)$ where 
\begin{align}
  W_n^{SP}(\mu_\delta^\ast, \alpha_\delta^\ast) = W_{1,n}^E(\mu_\delta^\ast) + W_{2,n}(\alpha_\delta^\ast) \label{eq:LRT_SP_total}
\end{align}
By the non-parametric Wilk's theorem [@owen1988empirical] it follows that $W_n^{SP}(0, 0) \sim \chi^2_2$ asymptotically.

@@

To assess the power we vary the sample size between 50 and 350 observations in each group with increments of 25, and for each scenario we perform 25,000 simulations and estimate the power as the proportion of rejected null hypotheses at the 5\% level. We compare the power of both our parametric and semi-parametric likelihood ratio tests to the two-sample t-test and the Wilcoxon test. Figure \ref{fig:powerCurves} shows the estimated power functions. The results are also presented in table form in the supplementary material.

@@

In addition to the power simulations we also performed a simulation study for the type I error. We considered two scenarios similar to scenarios 1 and 4 in Table \ref{tab:simulation} under a null-hypothesis of no treatment effect. Specifically, we considered $A_i \sim \Bernoulli(\pi)$, $Y_i \mid A_i = 1 \sim N(3,1)$ and $A_i \sim \Bernoulli(\pi)$, $Y_i \mid A_i = 1 \sim \Beta(1, 1)$. We estimated the type I error by simulations for different sample sizes and $\pi \in \left\{0.2, 0.4, 0.6, 0.8\right\}$. The results are presented in the supplementary material.

@@

The COVID-STEROID2 trial was a multi-center study comparing daily dose 6 mg (low) of dexamethasone for up to 10 days vs. 12 mg (high) in patients with severe and critical COVID-19. The full description of the study including all results can be found in @10.1001/jama.2021.18295. The primary outcome of the trial was days alive and without out use of life-support in the period from randomization to day 28. In other words, that outcome is an integer between 0 and 28 with a substantial peak at day 0 corresponding to the atom $\mathcal{E}$ coming from the patients who are never taken off life-support systems. The analyses in the main publications (@10.1001/jama.2021.18295) were conducted using the test proposed in this paper. In the JAMA publication the test is being referred to as the Kryger Jensen and Lange test. In this section we re-analyze the sub-group consisting of patients in need of invasive mechanical ventilation at baseline (n = 206). These can generally be taken to be the most severely ill patients.

@@

```{r, echo = FALSE, fig.width = 8, fig.height = 3, fig.cap="\\label{fig:application}Histograms of days alive without life support from the The COVID-STEROID2 trial stratified by treatment dosis (left, middle) and simultaneous confidence regions based on the semi-parametric likelihood ratio test (right).", fig.pos="htbp"}
par(mfrow=c(1,3), mgp=c(2,1,0), mar=c(3, 3, 2, 0))
hist(subset(d, d$allocation)$dawols28, xlim=c(0,30),
     main="Low dose", ylim=c(0,55), 10, col="gray70", border="gray90",
     cex.axis=1.2, cex.main=1.5, cex.lab=1.3, ylab="",
     xlab="Days alive without life support")
hist(subset(d, !d$allocation)$dawols28, 10, xlim=c(0,30),
     main="High dose", ylim = c(0,55), col="gray70", border="gray90",
     cex.axis=1.2, cex.main=1.5, cex.lab=1.3, ylab="",
     xlab="Days alive without life support")
image(sCI$muDelta, sCI$logORdelta, sCI$surface,
     col = rev(fields::tim.colors(256)), useRaster = TRUE,
     xlab="Difference in means",
     ylab="log OR",
     cex.axis=1.2, cex.main=1.5, cex.lab=1.3)
points(m_SPLRT$muDelta, log(m_SPLRT$alphaDelta), pch = 19, cex = 1.5)
contour(sCI$muDelta, sCI$logORdelta, sCI$surface, add = TRUE, 
        levels = stats::qchisq(0.99, 2), lwd = 1, labels = 0.99, lty=1)
contour(sCI$muDelta, sCI$logORdelta, sCI$surface, add = TRUE, 
        levels = stats::qchisq(0.95, 2), lwd = 1, labels = 0.95, lty=1)
contour(sCI$muDelta, sCI$logORdelta, sCI$surface, add = TRUE, 
        levels = stats::qchisq(0.8, 2), lwd = 1, labels = 0.8, lty=1)
contour(sCI$muDelta, sCI$logORdelta, sCI$surface, add = TRUE, 
        levels = stats::qchisq(0.5, 2), lwd = 1, labels = 0.5, lty=1)
title("Confidence regions", cex.main=1.5)
abline(v = 0, lty=2)
abline(h = 0, lty=2)
```

Comparing the groups by a Wilcoxon test yields a p-value of $0.242$ hence leading to a non-significant conclusion at the pre-specified 5\% level. Using our proposed parametric likelihood ratio test yields a p-value of $0.019$ and therefore leading to a rejection of the null-hypothesis. From the histograms it could be questioned whether the continuous part of the distributions are normal distributed. Accordingly, we also apply our proposed semi-parametric version of the likelihood ratio test. This yields a p-value of $0.021$. In conclusion, our proposed method is able to detect the difference between the groups. Further and in contrast to the Wilcoxon test, our method also provides a way to interpret the difference between the groups. This is illustrated in the right panel of Figure \ref{fig:application} showing simultaneous confidence regions for the two effect parameters. In this case it is clear that the primary effect of the intervention is not on the discrete component (i.e., the proportion of patients who never get off life support). Instead, the effect appears to be on the continuous part and is between half a day and seven days.

@@

\textbf{[TODO: check verb tense]}
